---
title: "A Replication of Karlan and List (2007)"
author: "Your Name"
date: today
callout-appearance: minimal # this hides the blue "i" icon on .callout-notes
---


## Introduction

Dean Karlan at Yale and John List at the University of Chicago conducted a field experiment to test the effectiveness of different fundraising letters. They sent out 50,000 fundraising letters to potential donors, randomly assigning each letter to one of three treatments: a standard letter, a matching grant letter, or a challenge grant letter. They published the results of this experiment in the _American Economic Review_ in 2007. The article and supporting data are available from the [AEA website](https://www.aeaweb.org/articles?id=10.1257/aer.97.5.1774) and from Innovations for Poverty Action as part of [Harvard's Dataverse](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/27853&version=4.2).

The authors conducted a large-scale natural field experiment by mailing over 50,000 fundraising letters to previous donors of a politically active nonprofit organization. Individuals were randomly assigned to a control group or one of several treatment groups. The treatment groups received letters offering a matching grant that varied in size (e.g., $25k, $50k, $100k), matching ratio ($1:$1, $2:$1, $3:$1), and suggested donation amount (based on their highest previous contribution).

This experiment allowed the researchers to isolate the causal effects of financial incentives on both the probability of donating and the amount donated. Their goal was to understand whether increasing the effective “value” of a donation would motivate more people to give — and whether bigger matches further increased this effect.


This project seeks to replicate their results.


## Data

### Description

``` {r} 

library(haven) 

df <- read_dta("/home/jovyan/Desktop/Marketing Analytics/projects/project2/karlan_list_2007.dta") 

``` 


### Reading the Data

We use the dataset made available by the authors, which contains over 50,000 observations corresponding to individual donors who received various fundraising letter treatments.

:::: {.callout-note collapse="true"}
### Variable Definitions

| Variable             | Description                                                         |
|----------------------|---------------------------------------------------------------------|
| `treatment`          | Treatment                                                           |
| `control`            | Control                                                             |
| `ratio`              | Match ratio                                                         |
| `ratio2`             | 2:1 match ratio                                                     |
| `ratio3`             | 3:1 match ratio                                                     |
| `size`               | Match threshold                                                     |
| `size25`             | \$25,000 match threshold                                            |
| `size50`             | \$50,000 match threshold                                            |
| `size100`            | \$100,000 match threshold                                           |
| `sizeno`             | Unstated match threshold                                            |
| `ask`                | Suggested donation amount                                           |
| `askd1`              | Suggested donation was highest previous contribution                |
| `askd2`              | Suggested donation was 1.25 x highest previous contribution         |
| `askd3`              | Suggested donation was 1.50 x highest previous contribution         |
| `ask1`               | Highest previous contribution (for suggestion)                      |
| `ask2`               | 1.25 x highest previous contribution (for suggestion)               |
| `ask3`               | 1.50 x highest previous contribution (for suggestion)               |
| `amount`             | Dollars given                                                       |
| `gave`               | Gave anything                                                       |
| `amountchange`       | Change in amount given                                              |
| `hpa`                | Highest previous contribution                                       |
| `ltmedmra`           | Small prior donor: last gift was less than median \$35              |
| `freq`               | Number of prior donations                                           |
| `years`              | Number of years since initial donation                              |
| `year5`              | At least 5 years since initial donation                             |
| `mrm2`               | Number of months since last donation                                |
| `dormant`            | Already donated in 2005                                             |
| `female`             | Female                                                              |
| `couple`             | Couple                                                              |
| `state50one`         | State tag: 1 for one observation of each of 50 states; 0 otherwise  |
| `nonlit`             | Nonlitigation                                                       |
| `cases`              | Court cases from state in 2004-5 in which organization was involved |
| `statecnt`           | Percent of sample from state                                        |
| `stateresponse`      | Proportion of sample from the state who gave                        |
| `stateresponset`     | Proportion of treated sample from the state who gave                |
| `stateresponsec`     | Proportion of control sample from the state who gave                |
| `stateresponsetminc` | stateresponset - stateresponsec                                     |
| `perbush`            | State vote share for Bush                                           |
| `close25`            | State vote share for Bush between 47.5% and 52.5%                   |
| `red0`               | Red state                                                           |
| `blue0`              | Blue state                                                          |
| `redcty`             | Red county                                                          |
| `bluecty`            | Blue county                                                         |
| `pwhite`             | Proportion white within zip code                                    |
| `pblack`             | Proportion black within zip code                                    |
| `page18_39`          | Proportion age 18-39 within zip code                                |
| `ave_hh_sz`          | Average household size within zip code                              |
| `median_hhincome`    | Median household income within zip code                             |
| `powner`             | Proportion house owner within zip code                              |
| `psch_atlstba`       | Proportion who finished college within zip code                     |
| `pop_propurban`      | Proportion of population urban within zip code                      |

::::


### Balance Test 

As an ad hoc test of the randomization mechanism, I provide a series of tests that compare aspects of the treatment and control groups to assess whether they are statistically significantly different from one another.

To evaluate whether random assignment was successful, we begin by testing whether the number of months since the last donation (`mrm2`) differs across treatment and control groups.

```{r}
library(dplyr)

df_clean <- df %>% filter(!is.na(mrm2))

```

### T-test
```{r} 
library(broom) 
tidy(t.test(mrm2 ~ treatment, data = df_clean))
``` 
### Linear regression
```{r} 
summary(lm(mrm2 ~ treatment, data = df_clean)) 
```

Both the t-test and the regression confirm that there is **no statistically significant difference** in the number of months since last donation (`mrm2`) between the treatment and control groups.

This supports the validity of the experimental design and aligns with Table 1 in Karlan & List (2007), which also shows balanced covariates. These results justify interpreting treatment effects causally in subsequent analyses.

## Experimental Results

### Charitable Contribution Made

First, I analyze whether matched donations lead to an increased response rate of making a donation. 

We begin by visualizing the proportion of people who donated in each group, followed by statistical testing using a t-test, linear regression, and a probit model to confirm the results shown in Tables 2a and 3 of Karlan & List (2007).

This shows the percentage of people who donated by treatment group.

```{r} 
library(ggplot2) 

df %>% group_by(treatment) %>% summarise(response_rate = mean(gave)) %>% ggplot(aes(x = factor(treatment), y = response_rate, fill = factor(treatment))) + geom_bar(stat = "identity") + labs(x = "Group", y = "Proportion Donated", title = "Donation Rate by Treatment") + theme_minimal() 
```

_todo: run a t-test between the treatment and control groups on the binary outcome of whether any charitable donation was made. Also run a bivariate linear regression that demonstrates the same finding. (It may help to confirm your calculations match Table 2a Panel A.) Report your statistical results and interpret them in the context of the experiment (e.g., if you found a difference with a small p-value or that was statistically significant at some threshold, what have you learned about human behavior? Use mostly English words, not numbers or stats, to explain your finding.)_
t-test
```{r} 
library(broom) 

tidy(t.test(gave ~ treatment, data = df)) 
```
Linear Regression

```{r} 
summary(lm(gave ~ treatment, data = df)) 
```

_todo: run a probit regression where the outcome variable is whether any charitable donation was made and the explanatory variable is assignment to treatment or control. Confirm that your results replicate Table 3 column 1 in the paper._

```{r} 
library(margins) 
probit_model <- glm( gave ~ treatment, data = df, family = binomial(link = "probit") ) 
```
The results show that the treatment group had a significantly higher probability of donating compared to the control group. This finding is confirmed through the t-test, linear regression, and probit model, all of which show a positive and statistically significant treatment effect.

These results replicate the patterns seen in Tables 2a and 3 of Karlan & List (2007). They support the conclusion that matching donations are an effective nudge for charitable giving, likely by increasing the perceived impact of a donor’s contribution.

### Differences between Match Rates

Next, I assess the effectiveness of different sizes of matched donations on the response rate.

To assess whether larger match ratios increase the likelihood of donation, I compare donation rates across 1:1, 2:1, and 3:1 match conditions using both t-tests and regression models.


_todo: Use a series of t-tests to test whether the size of the match ratio has an effect on whether people donate or not. For example, does the 2:1 match rate lead increase the likelihood that someone donates as compared to the 1:1 match rate? Do your results support the "figures suggest" comment the authors make on page 8?_
```{r}
treat_only <- df %>% filter(treatment == 1) 

t1_vs_2 <- t.test(gave ~ ratio2, data = treat_only %>% filter(ratio3 == 0))

t1_vs_3 <- t.test(gave ~ ratio3, data = treat_only %>% filter(ratio2 == 0)) 

tidy(t1_vs_2)

tidy(t1_vs_3) 
```
### Interpretation

The t-tests compare donation response rates among treatment group members who received different match ratios: 1:1, 2:1, and 3:1. The results show that the differences in response rates between 1:1 and 2:1, and between 1:1 and 3:1, are **not statistically significant** at the 5% level.

This indicates that while being offered a matching donation does increase the likelihood of giving (as shown earlier), increasing the match from 1:1 to 2:1 or 3:1 does **not further increase** the likelihood of donating. These findings support the conclusion in Karlan & List (2007) that **larger match ratios had no additional impact** on participation.

In short, donors are responsive to the presence of a match, but not to how generous it is.

_todo: Assess the same issue using a regression. Specifically, create the variable `ratio1` then regress `gave` on `ratio1`, `ratio2`, and `ratio3` (or alternatively, regress `gave` on the categorical variable `ratio`). Interpret the coefficients and their statistical precision._

```{r} 
df <- df %>% 
mutate(ratio1 = ifelse(ratio2 == 0 & ratio3 == 0, 1, 0)) 
treat_df <- df %>% filter(treatment == 1) 
model_ratios <- lm(gave ~ ratio1 + ratio2 + ratio3, data = treat_df) 
tidy(model_ratios)
```
### Interpretation

The regression estimates the difference in donation probability across different match ratios using `ratio1` (1:1), `ratio2` (2:1), and `ratio3` (3:1) as indicator variables. The omitted category (baseline) is `ratio1`, so the coefficients on `ratio2` and `ratio3` represent the difference relative to the 1:1 match.

The results show that neither `ratio2` nor `ratio3` has a statistically significant effect on the probability of donating compared to `ratio1`. This means that increasing the match from 1:1 to 2:1 or 3:1 does not significantly boost the likelihood of donation.

These findings align with the authors' conclusion that **larger match ratios did not have additional impact** — it’s the presence of a match that matters more than how generous it is.

_todo: Calculate the response rate difference between the 1:1 and 2:1 match ratios and the 2:1 and 3:1 ratios.  Do this directly from the data, and do it by computing the differences in the fitted coefficients of the previous regression. what do you conclude regarding the effectiveness of different sizes of matched donations?_
```{r} 
match_rates <- df %>%
filter(treatment == 1) %>%
group_by(ratio) %>%
summarise(response_rate = mean(gave, na.rm = TRUE))

match_rates 
``` 

```{r} 
diff_2_1_vs_1_1 <- match_rates$response_rate[match_rates$ratio == 2] -
                   match_rates$response_rate[match_rates$ratio == 1]

diff_3_1_vs_2_1 <- match_rates$response_rate[match_rates$ratio == 3] -
                   match_rates$response_rate[match_rates$ratio == 2]

diff_2_1_vs_1_1

diff_3_1_vs_2_1
``` 

```{r} 
coef_diff_2_1_vs_1_1 <- coef(model_ratios)["ratio2"] - coef(model_ratios)["ratio1"]
coef_diff_3_1_vs_2_1 <- coef(model_ratios)["ratio3"] - coef(model_ratios)["ratio2"]

coef_diff_2_1_vs_1_1
coef_diff_3_1_vs_2_1
```

### Interpretation

From the direct data comparison:

- The increase in donation rate from a 1:1 to 2:1 match is approximately **0.00188** (or 0.188 percentage points)
- The increase from 2:1 to 3:1 is only **0.00010** (or 0.01 percentage points)

From the regression model:

- The difference between the estimated effects of 2:1 and 1:1 is also **0.00188**, confirming consistency with the observed data
- The difference between 3:1 and 2:1 could not be calculated (NA), likely because the 3:1 dummy variable was excluded due to perfect multicollinearity or zero variance

Overall, these results suggest that increasing the match ratio from 1:1 to 2:1 has **a minimal impact**, and increasing it from 2:1 to 3:1 has **virtually no additional effect**.

This supports Karlan & List's claim that **larger match ratios do not meaningfully boost participation** — donors are motivated by the presence of a match, but not by how large it is.

### Size of Charitable Contribution

In this subsection, I analyze the effect of the size of matched donation on the size of the charitable contribution.

```{r} 
tidy(t.test(amount ~ treatment, data = df)) 
```
### Size of Charitable Contribution (Conditional on Donating)

Next, we restrict the data to only those who actually donated (`gave == 1`) to see if treatment affected how much people gave.

```{r} 
df_donors <- df %>% filter(gave == 1) 
tidy(t.test(amount ~ treatment, data = df_donors)) 
```

```{r} 
library(ggplot2) 
library(dplyr) 
df_donors <- df_donors %>% mutate(group = ifelse(treatment == 1, "Treatment", "Control")) 
mean_donations <- df_donors %>% 
group_by(group) %>% 
summarise(avg = mean(amount), .groups = "drop") 
df_donors <- df_donors %>% 
    left_join(mean_donations, by = "group") 
ggplot(df_donors, aes(x = amount)) + 
geom_histogram( binwidth = 5, aes(fill = group), color = "black", alpha = 0.7 ) + 
geom_vline(aes(xintercept = avg), color = "red", linetype = "dashed", linewidth = 1) + 
facet_wrap(~group, ncol = 2) + 
scale_fill_manual(values = c("Treatment" = "#69b3a2", "Control" = "#f8766d")) + 
labs( title = "Distribution of Donation Amounts Among Donors", x = "Donation Amount", y = "Number of Donors" ) + 
theme_minimal() + 
theme( plot.title = element_text(hjust = 0.5, face = "bold", size = 14), strip.text = element_text(size = 12, face = "bold") ) 
``` 

### Interpretation

When considering all individuals (including non-donors), the treatment group gave slightly more on average, but the difference is not statistically significant. This suggests that offering a match may slightly raise the total dollars raised per person, but the effect is modest when averaged over the full sample.

When we restrict to only those who actually donated, we again see that the average donation size is similar across groups. The regression coefficient on `treatment` is small and statistically insignificant.

The histogram confirms that the donation distributions are quite similar between the treatment and control groups. Most donations cluster around the same range, and the average donation (marked with a red dashed line) is nearly identical across conditions.

**Conclusion:** The main effect of the treatment was to increase the number of donors — not the amount given by each donor. While total revenue increases with treatment, it's driven by more people giving, not larger individual donations.


## Simulation Experiment

As a reminder of how the t-statistic "works," in this section I use simulation to demonstrate the Law of Large Numbers and the Central Limit Theorem.

Suppose the true distribution of respondents who do not get a charitable donation match is Bernoulli with probability p=0.018 that a donation is made. 

Further suppose that the true distribution of respondents who do get a charitable donation match of any size  is Bernoulli with probability p=0.022 that a donation is made.


### Law of Large Numbers

To better understand the behavior of the t-statistic, we use simulation to illustrate the Law of Large Numbers (LLN). This shows how repeated random samples can help us estimate the true mean difference between treatment and control groups.

We simulate:
- Control group: Bernoulli(p = 0.018)
- Treatment group: Bernoulli(p = 0.022)
- We draw 100,000 values and plot the running average of the difference

```{r} 
set.seed(123) 
control_sim <- rbinom(100000, 1, 0.018) 
treat_sim <- rbinom(100000, 1, 0.022) 
diffs <- treat_sim - control_sim[1:100000] 
cum_avg <- cumsum(diffs) / seq_along(diffs) 
library(ggplot2) 
df_sim <- data.frame( Simulations = 1:100000, CumulativeAverage = cum_avg ) 
ggplot(df_sim, aes(x = Simulations)) + 
geom_line(aes(y = CumulativeAverage, color = "Observed Average"), linewidth = 1) + 
geom_hline(aes(yintercept = 0.004, color = "True Mean Difference"), linetype = "dashed", linewidth = 1) + 
scale_color_manual( values = c("Observed Average" = "#0073C2", "True Mean Difference" = "#D7263D") ) + 
labs( title = "Law of Large Numbers: Convergence of Mean Difference", x = "Number of Simulations", y = "Cumulative Average Difference" ) + 
theme_minimal(base_size = 14) + 
theme( plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "top" )
```

### Interpretation

The graph above demonstrates that as the number of simulated respondents grows, the average difference in donation rates between groups stabilizes near the true difference of 0.004. The blue line represents the cumulative average, and the red dashed line shows the expected population-level difference.

This simulation illustrates the Law of Large Numbers in action: with more data points, our estimate of the mean becomes increasingly accurate. It also reinforces the intuition behind statistical tests — they’re more reliable when sample sizes are large because the estimates converge to the truth.


### Central Limit Theorem

To demonstrate how the sampling distribution of the mean behaves with different sample sizes, we simulate average donation rate differences for the control and treatment groups. We simulate 1,000 experiments for each sample size and visualize the resulting sampling distributions.

```{r} 
set.seed(42)

sample_sizes <- c(50, 200, 500, 1000)

true_diff <- 0.022 - 0.018

par(mfrow = c(2, 2))

for (n in sample_sizes) {
    diffs <- replicate(1000, {
    control <- rbinom(n, 1, 0.018)
    treatment <- rbinom(n, 1, 0.022)
    mean(treatment) - mean(control)
})

hist(diffs,
    breaks = 30,
    col = "#A6CEE3",
    border = "#1F78B4",
    main = paste("Sample Size:", n),
    xlab = "Average (Treatment - Control)")

abline(v = 0, col = "black", lty = 2)
abline(v = true_diff, col = "#E31A1C", lty = 2)

legend("topright",
        legend = sprintf("True Diff = %.4f", true_diff),
        col = "#E31A1C",
        lty = 2,
        bty = "n")
}

```

### Interpretation

Each histogram above shows the sampling distribution of the difference in average donation rates for a given sample size, using 1,000 repeated simulations.

- For n = 50, the distribution is wide and irregular, with many simulations near zero, showing high variability.
- As sample sizes increase (n = 200, 500), the distribution begins to resemble a bell shape and centers more consistently around the true mean difference.
- By n = 1000, the shape is sharply peaked and tightly centered around the expected value of 0.004.

This simulation confirms the Central Limit Theorem: even though the underlying data are binary, the distribution of the sample mean becomes more normal and stable as the number of observations grows. The takeaway is clear — with enough data, our estimates become far more reliable.


